\hypertarget{bibliography-working-notes}{%
\section{8. Bibliography (working
notes)}\label{bibliography-working-notes}}

\begin{quote}
The authoritative file is \texttt{latex/bibliography.bib}. This Markdown
file collects raw references and editorial notes during writing. Entries
are migrated to BibTeX as they are cited in the main text.
\end{quote}

A subset of the references below is \textbf{shared with the companion
thesis} {[}Motje, 2026{]}. The two theses are written to be read
together and deliberately use a unified set of foundational citations on
knowledge-management theory, ICT-for-development, community networks,
and core network protocols. Where a reference is shared, it is marked
\textbf{{[}shared{]}} below.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{companion-thesis}{%
\subsection{Companion thesis}\label{companion-thesis}}

\begin{itemize}
\tightlist
\item
  \textbf{{[}Motje, 2026{]}} Motje, J. \emph{Software services for
  community networks: a reproducible, open field guide built with the
  AUCOOP association.} Master Thesis, Universitat Politècnica de
  Catalunya (UPC), Master in Telecommunications Engineering (MET), 2026.
  Companion thesis to this work; covers the software/services
  work-stream (virtualisation, identity, file services, monitoring, VPN,
  graceful shutdown).
\end{itemize}

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{knowledge-management-and-learning-organisations-shared-backbone}{%
\subsection{Knowledge management and learning organisations {[}shared
backbone{]}}\label{knowledge-management-and-learning-organisations-shared-backbone}}

\begin{itemize}
\tightlist
\item
  Nonaka, I. and Takeuchi, H. \emph{The Knowledge-Creating Company.}
  Oxford University Press, 1995. \textbf{{[}shared{]}}
\item
  Wenger, E. \emph{Communities of Practice: Learning, Meaning, and
  Identity.} Cambridge University Press, 1998. \textbf{{[}shared{]}}
\item
  Davenport, T. H. and Prusak, L. \emph{Working Knowledge: How
  Organizations Manage What They Know.} Harvard Business Press, 1998.
  \textbf{{[}shared{]}}
\item
  Cuskelly, G., Hoye, R., and Auld, C. \emph{Working with Volunteers in
  Sport: Theory and Practice.} Routledge, 2006. \textbf{{[}shared{]}}
  --- volunteer-rotation and continuity literature.
\item
  Hager, M. and Brudney, J. \emph{Volunteer Management Practices and
  Retention of Volunteers.} The Urban Institute, 2004.
  \textbf{{[}shared{]}}
\end{itemize}

\hypertarget{ict-for-development-and-the-digital-divide-shared-backbone}{%
\subsection{ICT for development and the digital divide {[}shared
backbone{]}}\label{ict-for-development-and-the-digital-divide-shared-backbone}}

\begin{itemize}
\tightlist
\item
  ITU. \emph{Measuring digital development: Facts and figures} (latest
  annual report). International Telecommunication Union, Geneva.
  \textbf{{[}shared{]}}
\item
  Heeks, R. \emph{Information and Communication Technology for
  Development (ICT4D).} Routledge, 2017. \textbf{{[}shared{]}}
\item
  Toyama, K. \emph{Geek Heresy: Rescuing Social Change from the Cult of
  Technology.} PublicAffairs, 2015. \textbf{{[}shared{]}}
\item
  van Dijk, J. \emph{The Digital Divide.} Polity, 2020.
  \textbf{{[}shared{]}}
\end{itemize}

\hypertarget{community-networks}{%
\subsection{Community networks}\label{community-networks}}

\begin{itemize}
\tightlist
\item
  Belli, L. (ed.). \emph{Community Networks: the Internet by the People,
  for the People.} FGV Direito Rio, 2017. \textbf{{[}shared{]}}
\item
  APC. \emph{Local Access Networks: can the unconnected connect
  themselves?} Association for Progressive Communications.
  \textbf{{[}shared{]}}
\item
  Guifi.net Foundation. \url{https://guifi.net/}
\item
  AlterMundi. \url{https://altermundi.net/}
\end{itemize}

\hypertarget{refurbishment-and-circular-economy}{%
\subsection{Refurbishment and circular
economy}\label{refurbishment-and-circular-economy}}

\begin{itemize}
\tightlist
\item
  Labdoo project. \url{https://www.labdoo.org/}
\item
  eReuse / DeviceHub. \url{https://ereuse.org/}
\item
  WEEE Directive 2012/19/EU. \emph{Waste Electrical and Electronic
  Equipment.} Official Journal of the European Union, 2012.
\end{itemize}

\hypertarget{openwrt-mesh-and-wi-fi}{%
\subsection{OpenWrt, mesh, and Wi-Fi}\label{openwrt-mesh-and-wi-fi}}

\begin{itemize}
\tightlist
\item
  OpenWrt Project. \emph{Documentation.} \url{https://openwrt.org/docs/}
\item
  IEEE Std 802.11s-2011. \emph{Mesh Networking Amendment to IEEE Std
  802.11.} IEEE, 2011.
\item
  B.A.T.M.A.N. advanced. \url{https://www.open-mesh.org/}
\end{itemize}

\hypertarget{core-network-protocols-shared-backbone}{%
\subsection{Core network protocols {[}shared
backbone{]}}\label{core-network-protocols-shared-backbone}}

\begin{itemize}
\tightlist
\item
  Rekhter, Y. et al.~\emph{RFC 1918 --- Address Allocation for Private
  Internets.} IETF, 1996. \textbf{{[}shared{]}}
\item
  Rigney, C. et al.~\emph{RFC 2865 --- Remote Authentication Dial In
  User Service (RADIUS).} IETF, 2000. \textbf{{[}shared{]}} (cited
  primarily by {[}Motje, 2026{]}; appears here for boundary-services
  completeness.)
\item
  Droms, R. \emph{RFC 2131 --- Dynamic Host Configuration Protocol.}
  IETF,

  \begin{enumerate}
  \def\labelenumi{\arabic{enumi}.}
  \setcounter{enumi}{1996}
  \tightlist
  \item
    \textbf{{[}shared{]}}
  \end{enumerate}
\item
  Mockapetris, P. \emph{RFC 1035 --- Domain Names --- Implementation and
  Specification.} IETF, 1987. \textbf{{[}shared{]}}
\end{itemize}

\hypertarget{pxe-network-boot-and-imaging}{%
\subsection{PXE, network boot, and
imaging}\label{pxe-network-boot-and-imaging}}

\begin{itemize}
\tightlist
\item
  Clonezilla project. \url{https://clonezilla.org/}
\item
  DRBL --- Diskless Remote Boot in Linux. \url{https://drbl.org/}
\item
  Free Software Foundation. \emph{GNU GRUB Manual --- Network booting.}
  \url{https://www.gnu.org/software/grub/manual/grub/html_node/Network.html}
\item
  Intel Corporation. \emph{Preboot Execution Environment (PXE)
  Specification, Version 2.1.} 1999.
\item
  ISC. \emph{isc-dhcp-server documentation.}
  \url{https://kb.isc.org/docs/isc-dhcp-44-manual-pages-dhcpdconf}
\item
  tftpd-hpa manual. \url{https://manpages.debian.org/tftpd-hpa}
\end{itemize}

\hypertarget{linux-distributions}{%
\subsection{Linux distributions}\label{linux-distributions}}

\begin{itemize}
\tightlist
\item
  Linux Mint Project. \url{https://linuxmint.com/}
\item
  Ubuntu LTS release notes (relevant version for Mint 22.3 base).
\item
  OnlyOffice. \url{https://www.onlyoffice.com/}
\end{itemize}

\hypertarget{power-systems-and-resilience}{%
\subsection{Power systems and
resilience}\label{power-systems-and-resilience}}

\begin{itemize}
\tightlist
\item
  IEEE Std 450-2010. \emph{Recommended Practice for Maintenance,
  Testing, and Replacement of Vented Lead-Acid Batteries for Stationary
  Applications.}
\item
  ITU-T L.1200 series. \emph{Power feeding system for ICT equipment.}
\item
  Network UPS Tools (NUT). \url{https://networkupstools.org/} (cited
  primarily by {[}Motje, 2026{]}.)
\end{itemize}

\hypertarget{living-documentation-and-docs-as-code}{%
\subsection{Living documentation and
docs-as-code}\label{living-documentation-and-docs-as-code}}

\begin{itemize}
\tightlist
\item
  Gentle, A. \emph{Docs Like Code.} 3rd ed., 2022.
\item
  GitLab. \emph{The GitLab Handbook.}
  \url{https://about.gitlab.com/handbook/}
\item
  Arch Linux. \emph{ArchWiki contributing guidelines.}
\item
  Diátaxis documentation framework. \url{https://diataxis.fr/} --- the
  four-mode (tutorials/how-to/reference/explanation) framework that
  inspired the Ch2 (story) ↔ Ch3 (recipe) split of §3.C.2.
\end{itemize}

\hypertarget{tooling-for-the-handbook}{%
\subsection{Tooling for the handbook}\label{tooling-for-the-handbook}}

\begin{itemize}
\tightlist
\item
  Zensical static-site generator (used for the handbook build).
\item
  Material for MkDocs theme.
  \url{https://squidfunk.github.io/mkdocs-material/}
\item
  Mermaid.js. \url{https://mermaid.js.org/}
\item
  OpenCode editor and subagent framework. \url{https://opencode.ai/}
\end{itemize}

\hypertarget{this-work}{%
\subsection{This work}\label{this-work}}

\begin{itemize}
\tightlist
\item
  AUCOOP. \emph{Community Network Handbook.} Branch
  \texttt{dev\_mj\_thesis} @ commit
  \texttt{a5fc80b26368fe6cf1475994a12cc8f1967b84f0}.
  \url{https://github.com/aucoop/Community-Network-Handbook}
\item
  Jover, M. \emph{Master thesis source repository.}
  \url{https://github.com/Maria-Jover/master-thesis-ict-deployment}
\end{itemize}

\hypertarget{case-study-namibia-2026}{%
\subsection{§4 --- case study (Namibia
2026)}\label{case-study-namibia-2026}}

\begin{itemize}
\tightlist
\item
  AUCOOP. \emph{Gochas project diary.}
  \url{https://aucoop.upc.edu/gochas-namibia/}
\item
  Foundawtion (local partner).
  \url{https://foundawtion.org/en/home-english}
\item
  N Mutschuana Primary School, Gochas, Hardap Region, Namibia
  (deployment site, March 2026).
\item
  UPC CCD (Centre de Cooperació per al Desenvolupament). Funding source,
  5 600 €.
\item
  NexTReT. Equipment donor (3 laptops, 2 mini-PCs).
\end{itemize}

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{todo}{%
\subsection{TODO}\label{todo}}

\begin{itemize}
\tightlist
\item
  Migrate every reference above to BibTeX entries in
  \texttt{latex/bibliography.bib}.
\item
  Realign with {[}Motje, 2026{]} on the final shared subset and the
  exact edition/version of each shared reference.
\item
  Decide style: current template default is \texttt{unsrt}; consider
  \texttt{IEEEtran} for consistency with UPC MET conventions. Pending
  alignment with the companion thesis to keep the printed bibliography
  visually consistent across the two volumes.
\end{itemize}
